---
title: "proj2"
author: "Marco Pasin"
date: "26 January 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Assignment 2 - Exploratory Data Analysis

###Read files
```{r}
NEI <- readRDS("summarySCC_PM25.rds")
SCC <- readRDS("Source_Classification_Code.rds")
```



Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.
```{r}
totEmi<- tapply(NEI$Emissions, NEI$year, sum)
png('plot1.png')
barplot(totEmi, xlab = "year", ylab = "total emissions (ton)", 
        main = "Total Emissions per year",col = "blue")
dev.off()
plot(totEmi,type="l",xlab="year", ylab="emissions")
```

Have total emissions from PM2.5 decreased in the  Baltimore City, Maryland ( fips == 24510) from 1999 to 2008? Use the base plotting system to make a plot answering this question.
```{r}
bal <- subset(NEI, fips == "24510")
totEmiBal <- tapply(bal$Emissions, bal$year, sum)
png('plot2.png')
barplot(totEmiBal, xlab = "year", ylab = "total emissions (ton)", main = "Total Emission per year in Baltimore",col = "blue")
dev.off()
```

3. Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008? Use the ggplot2 plotting system to make a plot answer this question.
```{r}
bal <- subset(NEI, fips == "24510")
library(dplyr)
library(ggplot2)
EmiBalTyp<- bal%>%
  select(Emissions,type,year) %>%
  group_by(year,type) %>%
  summarize(Emissions=sum(Emissions))
EmiBalTyp$year<-as.factor(EmiBalTyp$year)
g<-ggplot(EmiBalTyp,aes(year,Emissions))
g+geom_bar(stat="identity")+facet_grid(.~type)+ggtitle("Total Emission per year in Baltimore by type of source")

```

4. Across the United States, how have emissions from coal combustion-related sources changed from 1999-2008?
```{r}
library(dplyr)
EmiTot<- NEI%>%
  select(SCC,Emissions,year) %>%
  group_by(year,SCC) %>%
  summarize(Emissions=sum(Emissions))
#do the merge
```

5. How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?

6.Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037").

7. Which city has seen greater changes over time in motor vehicle emissions?
