---
title: "Exploratory Data Analysis - Peer Assignment Week 1"
author: "Marco Pasin"
date: "08 gennaio 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis - Peer Assignment Week 1

### Read file and keep just necessary rows
```{r}
setwd("C:/Users/Marco/Dropbox/Coursera/Data Science Specialization - JHU/Exploratory Data Analysis")

houseConsumption<- read.table("household_power_consumption.txt",sep=";",header=TRUE)
head(houseConsumption)
str(houseConsumption)

library(dplyr)
houseConSub<- houseConsumption%>%
                        filter(Date=="1/2/2007"|Date=="2/2/2007")
summary(houseCon$Date)
```

### Convert to Date format and create new variable concatenating Date and Time
```{r}
houseConSub$Date<-as.Date(houseConSub$Date,"%d/%m/%Y")
houseConSub<- houseConSub%>%
                mutate(DateTime= as.POSIXct(paste(Date, Time)))
```

### Convert measure columns to numeric to be able to create plots later
```{r}
for(i in 3:9) {
    houseConSub[,i]<-as.numeric(as.character((houseConSub[,i])))
}
```


### Plot1
```{r}
hist(as.numeric(houseConSub$Global_active_power), col = "red", main = "Global Active Power",
     xlab = "Global Active Power (killowatts)", ylab = "Frequency")
```



## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
